---
title: "Activity 03: Summary Statistics"
author: "Ingrid Baduel"
date: "27 May 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

This activity explores **descriptive statistics and graphical representation** using a herpetofauna dataset from Palawan Island. The dataset contains species observation counts across multiple transect sites, categorized by habitat type and elevation.

---

## Part A: Read and Perform Descriptive Statistics

### A1. Load and Clean the Data

We begin by reading the CSV file and removing rows with missing values using `complete.cases()`.

```{r load-data}
herp <- read.csv("herpetozoa-033-095-s001.csv")

# Preview raw data before cleaning
head(herp)

# Remove rows with NAs
good <- complete.cases(herp)
herp <- herp[good, ]

# Confirm row count and preview cleaned data
nrow(herp)
head(herp)
```

---

### A2. Data Structure

Understanding the structure of our data — including variable names, types, and dimensions — is an essential first step before any analysis.

```{r structure}
str(herp)
```

---

### A3. Elevation Statistics

We extract the elevation column and compute its **mean**, **standard deviation**, and **range** to describe how elevation varies across sampling sites.

```{r elevation-stats}
elev <- herp[, "Elev"]

mean(elev)
sd(elev)
range(elev)
```

---

### A4. Species Counts

Species observation columns span columns 7 to 33. We count total observations per species using two methods — a `for` loop and `sapply()` — both producing the same result.

```{r species-counts}
spp <- herp[, 7:33]
head(spp)
ncol(spp)

## Method 1: for loop
spp_counts <- c()

for (i in 1:ncol(spp)) {
  spp_counts[i] <- sum(spp[, i])
}

names(spp_counts) <- names(spp)
print(spp_counts)

## Method 2: sapply (more concise)
spp_counts2 <- sapply(spp, sum)
spp_counts2
```

We then identify the species with the **most** and **fewest** total observations.

```{r most-fewest}
# Species with the most observations
spp_counts[spp_counts == range(spp_counts)[2]]

# Species with the fewest observations
spp_counts[spp_counts == range(spp_counts)[1]]
```

---

### A5. Mean Species Count per Habitat Type

We combine habitat type (`Htypes`) with the most observed species (`Phlo`) and the least observed (`Babu`), then compute the **mean count per habitat** using `split()` and `sapply()`.

```{r habitat-means}
df2 <- data.frame(Htypes = herp[, "Htypes"],
                  Phlo   = herp[, "Phlo"],
                  Babu   = herp[, "Babu"])
head(df2)

by_hab <- split(df2[, c("Phlo", "Babu")], df2$Htypes)
sapply(by_hab, function(x) colMeans(x))
```

> **Finding:** OGF (Old Growth Forest) has the highest mean for *Phlo*.

---

## Part B: Graphical Representation

### B1. Scatter Plot — Elevation vs. Total Species Count

Each point is colored by habitat type to reveal how species richness varies with elevation across different habitats.

```{r scatter-plot, fig.width=7, fig.height=5, fig.align='center'}
df3 <- data.frame(Elev     = herp[, "Elev"],
                  SPPTotal = herp[, "SPPTotal"],
                  Htypes   = herp[, "Htypes"])

hcols <- c("ASG" = "red", "CVT" = "blue", "ESG" = "darkgreen", "OGF" = "orange")

plot(df3$Elev, df3$SPPTotal,
     col  = hcols[df3$Htypes],
     pch  = 16,
     xlab = "Elevation (m)",
     ylab = "Total Species Count",
     main = "SPPTotal vs Elevation by Habitat")

legend("topright", legend = names(hcols), col = hcols, pch = 16)
```

> **Observation:** ASG, ESG, and CVT have sites with zero species. OGF always has species present.

---

### B2. Barplot — Total Species per Habitat Type

This barplot summarizes the **total species count** aggregated by habitat type using `tapply()`.

```{r barplot, fig.width=6, fig.height=4, fig.align='center'}
hab_total <- tapply(df3$SPPTotal, df3$Htypes, sum)

barplot(hab_total,
        xlab = "Habitat Type",
        ylab = "Total Species",
        main = "Total Species per Habitat Type",
        col  = c("red", "blue", "darkgreen", "orange"))
```

---

### B3. Histogram — Distribution of Total Species Counts

The histogram shows how total species counts are distributed across all sampling sites.

```{r histogram, fig.width=6, fig.height=4, fig.align='center'}
hist(df3$SPPTotal,
     xlab = "Total Species Count",
     ylab = "Frequency",
     main = "Histogram of Total Species Count")
```

---

## Part C: Interpretation

Based on the results, **OGF (Old Growth Forest)** has the highest total species count, while ASG, ESG, and CVT had multiple sites with zero observations. This suggests that disturbed habitats support less herpetofaunal activity.

The scatter plot also shows that higher species counts tend to occur at higher elevations, which mostly corresponds to OGF sites.

The species data is heavily skewed, with *Phlo* dominating at 377 observations while six species were only recorded once — a pattern common in ecological data. The histogram confirms this: most sites recorded very few or no species at all, with only a few high-count sites.

Overall, **habitat type and elevation together** seem to influence species richness, with old growth forest at higher elevations consistently supporting the most herpetofaunal diversity on Palawan Island.
