---
title: "Activity 03: Summary Statistics"
author: "Ingrid Baduel"
date: "01 June 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

## Part A: Read and Perform Descriptive Statistics

### A1. Load and Clean the Data


```{r load-data}
herp <- read.csv("herpetozoa-033-095-s001.csv")

# Preview raw data before cleaning
head(herp)

# Remove rows with NAs
good <- complete.cases(herp)
herp <- herp[good, ]

# Confirm row count and preview cleaned data
nrow(herp)
head(herp)
```

---

### A2. Data Structure


```{r structure}
str(herp)
```

---

### A3. Elevation Statistics


```{r elevation-stats}
elev <- herp[, "Elev"]

mean(elev)
sd(elev)
range(elev)
```

---

### A4. Species Counts


```{r species-counts}
spp <- herp[, 7:33]
head(spp)
ncol(spp)

## Method 1: for loop
spp_counts <- c()

for (i in 1:ncol(spp)) {
  spp_counts[i] <- sum(spp[, i])
}

names(spp_counts) <- names(spp)
print(spp_counts)

## Method 2: sapply (more concise)
spp_counts2 <- sapply(spp, sum)
spp_counts2
```

We then identify the species with the **most** and **fewest** total observations.

```{r most-fewest}
# Species with the most observations
spp_counts[spp_counts == range(spp_counts)[2]]

# Species with the fewest observations
spp_counts[spp_counts == range(spp_counts)[1]]
```

---

### A5. Mean Species Count per Habitat Type


```{r habitat-means}
df2 <- data.frame(Htypes = herp[, "Htypes"],
                  Phlo   = herp[, "Phlo"],
                  Babu   = herp[, "Babu"])
head(df2)

by_hab <- split(df2[, c("Phlo", "Babu")], df2$Htypes)
sapply(by_hab, function(x) colMeans(x))
```

> **Finding:** OGF (Old Growth Forest) has the highest mean for *Phlo*.

---

## Part B: Graphical Representation

### B1. Scatter Plot — Elevation vs. Total Species Count


```{r scatter-plot, fig.width=7, fig.height=5, fig.align='center'}
df3 <- data.frame(Elev     = herp[, "Elev"],
                  SPPTotal = herp[, "SPPTotal"],
                  Htypes   = herp[, "Htypes"])

hcols <- c("ASG" = "red", "CVT" = "blue", "ESG" = "darkgreen", "OGF" = "orange")

plot(df3$Elev, df3$SPPTotal,
     col  = hcols[df3$Htypes],
     pch  = 16,
     xlab = "Elevation (m)",
     ylab = "Total Species Count",
     main = "SPPTotal vs Elevation by Habitat")

legend("topright", legend = names(hcols), col = hcols, pch = 16)
```

> **Observation:** ASG, ESG, and CVT have sites with zero species. OGF always has species present.

---

### B2. Barplot — Total Species per Habitat Type

```{r barplot, fig.width=6, fig.height=4, fig.align='center'}
hab_total <- tapply(df3$SPPTotal, df3$Htypes, sum)

barplot(hab_total,
        xlab = "Habitat Type",
        ylab = "Total Species",
        main = "Total Species per Habitat Type",
        col  = c("red", "blue", "darkgreen", "orange"))
```

---

### B3. Histogram — Distribution of Total Species Counts


```{r histogram, fig.width=6, fig.height=4, fig.align='center'}
hist(df3$SPPTotal,
     xlab = "Total Species Count",
     ylab = "Frequency",
     main = "Histogram of Total Species Count")
```

---

## Part C: Interpretation

Based on the results, **OGF (Old Growth Forest)** has the highest total species count, while ASG, ESG, and CVT had multiple sites with zero observations. This suggests that disturbed habitats support less herpetofaunal activity.

The scatter plot also shows that higher species counts tend to occur at higher elevations, which mostly corresponds to OGF sites.

The species data is heavily skewed, with *Phlo* dominating at 377 observations while six species were only recorded once — a pattern common in ecological data. The histogram confirms this: most sites recorded very few or no species at all, with only a few high-count sites.

Overall, **habitat type and elevation together** seem to influence species richness, with old growth forest at higher elevations consistently supporting the most herpetofaunal diversity on Palawan Island.
