---
title: "Activity 04: Data Wrangling with Tidyverse"
author: "Ingrid Baduel"
date: "25 May 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

This activity applies **Tidyverse** tools to wrangle and explore a habitat dataset from Palawan Island. The dataset contains transect-level measurements of vegetation features across different habitat types. We use core Tidyverse functions — `select()`, `filter()`, `arrange()`, `group_by()`, `summarize()`, `mutate()`, and `pivot_wider()` — to extract insights from the data.

---

## Part A: Read and Filter

### A1. Install and Load Tidyverse

```{r load-packages}
# install.packages("tidyverse")  # Run once if not yet installed
library(tidyverse)
```

---

### A2. Read the CSV File

We use `read_csv()` from the `readr` package (part of Tidyverse) to load the habitat dataset.

```{r load-data}
habitat_df <- read_csv("habitat_types.csv")
```

---

### A3. Data Structure

`glimpse()` gives a compact, transposed view of the dataset — showing column names, types, and a preview of values.

```{r structure}
glimpse(habitat_df)
```

---

### A4. Select Relevant Columns

We keep only the four columns needed for our initial analysis: `Transect`, `Htypes`, `SmallTrees`, and `FloweringTrees`.

```{r select-cols}
sel_df <- habitat_df |>
  select(Transect, Htypes, SmallTrees, FloweringTrees)

print(sel_df)
```

---

### A5. Filter CVT Habitat and Sort by SmallTrees

We filter for **CVT (Cultivation Type)** sites only, then sort by `SmallTrees` in descending order to identify the transects with the most small trees.

```{r filter-cvt}
cvt_df <- sel_df |>
  filter(Htypes == "CVT") |>
  arrange(desc(SmallTrees))

print(cvt_df)
```

> **Finding:** IT2P7 and IT3P0 have the most small trees, each with a count of 10.

---

### A6. Mean Flowering Trees per Habitat Type

We group by habitat type and compute the **mean number of flowering trees** per group using `group_by()` and `summarize()`.

```{r mean-flowering}
mean_flowering <- sel_df |>
  group_by(Htypes) |>
  summarize(mean_FloweringTrees = mean(FloweringTrees, na.rm = TRUE))

print(mean_flowering)
```

> **Finding:** ESG (5.38) and ASG (5.26) have very similar means and do not differ significantly from each other. CVT has a notably higher mean (8.86).

---

### A7. Add a Derived Column: HerbGroundDiff

We expand our selected columns to include `HerbsCover` and `GroundCover`, then use `mutate()` to create a new column `HerbGroundDiff`, which is the difference between the two cover types.

```{r mutate-diff}
sel_df_mod <- habitat_df |>
  select(Transect, Htypes, SmallTrees, FloweringTrees, HerbsCover, GroundCover) |>
  mutate(HerbGroundDiff = HerbsCover - GroundCover)

print(sel_df_mod)
```

We then identify transects where herb and ground cover are equal (difference = 0).

```{r zero-diff}
sel_df_mod |>
  filter(HerbGroundDiff == 0) |>
  select(Transect, HerbGroundDiff)
```

> **Finding:** Transects IT1P5, IT2P5, and IT3P1 have no difference between herb and ground cover.

---

### A8. Reshape to Wide Format with `pivot_wider()`

We pivot the data to **wide format**, spreading habitat types across columns. Each row becomes a unique transect, and the values for `SmallTrees`, `FloweringTrees`, and `HerbGroundDiff` are spread into habitat-specific columns.

```{r pivot-wider}
sel_df_mod |>
  select(Transect, Htypes, SmallTrees, FloweringTrees, HerbGroundDiff) |>
  pivot_wider(names_from  = Htypes,
              values_from = c(SmallTrees, FloweringTrees, HerbGroundDiff)) ->
  wider_df

print(wider_df)
```

---

## Part B: Reflection

Tidyverse feels more readable than base R because of the **pipe operator** (`|>`): you can follow the logic step by step, like reading a sentence. Functions like `filter()`, `select()`, and `group_by()` are more intuitive than base R indexing with brackets.

The most challenging part was `pivot_wider()` — the `names_from` and `values_from` arguments took some getting used to. Overall, Tidyverse feels cleaner, especially for longer pipelines.
